Notebook: Predictors of frog survival following translocation

### 2022-05-30 doc/notebook/notebook_translocation.Rmd
Update text to reflect current status of analysis

-   site_id: site to which frogs were translocated
-   elevation
-   date
-   day
-   year
-   donor
-   order: first translocation to site = 0, subsequent translocation = 1
-   type: translocation conducted on foot versus by helicopter (not included)
-   shore: suitability of shore/bank habitat for frog overwintering
-   pit_tag_ref
-   survival: individual-level estimated survival 1 year following translocation, based on cmr surveys conducted for at least two years post-translocation
-   sex
-   length: at release
-   weight: at release
-   condition: index based on length and weight
-   swab_id
-   bd_load: at release

Maximum depth and surface area would seem useful to include, but 74976 is a stream/meadow habitat and as such depth and area at that site are likely not comparable to depth and area at lake habitats.

Frog condition might be problematic because size and residuals are correlated. Specifically, the largest frogs have the largest positive residuals from a weight:length regression line, even after log10 transformation. As such, length and condition are probably not independent.

```{r frog-length-weight-plot}
read_csv(here::here("data", "clean", "frog_translocation.csv")) %>% 
ggplot(aes(x = log10(length), y = log10(weight))) +
  geom_point() +
  geom_smooth(method = "lm")
plot
```

Another consideration when designing the analysis is that there are 12 records in which bd_load is null. Of these, 9 are from 70413 on 2013-07-15, likely due to a rush to get frogs packaged prior to helicopter arrival. 39 frogs were translocated on that date, so should probably drop the 9 records where bd_load = NA prior to running the analysis. If bd_load is not an important predictor, can run analysis using the full dataset.

@@

survival \~ elevation + shore + day + year + order + donor + sex + length + log(bdload) \| site_id

Group-level effects: In addition to the inclusion of site_id as a group-level effect, may need to include release_year, perhaps nested within site_id? Given that some sites only have a single release_year, will this cause problems?

Final steps before building model (in analysis Rmd):

-   Change to factor: order, day?, year?
-   Standardize variables?
-   Survival: Create 0/1 variable from probabilities. Change to factor?

## Results

To be added soon.

>>>>

* site_id: site to which frogs were translocated  
* elevation: water body surface elevation above sea level, in meters  
* shore: presence/absence of shore/bank habitat available to frogs in winter (estimated in late-summer)  
* snow_t: snow water content (measured on April 1) in year of translocation  
* snow_t1: snow water content (measured on April 1) in year following translocation  
* date: date frogs released at recipient site  
* year: year frogs released at recipient site  
* day: day of the year on which frogs were released  
* order: order of translocations to a site, from 1-4. Re-coded as indicator variable (variable name = "first"): 1 = first translocation to site; 0 = subsequent translocation  
* donor: site_id from which frogs were collected for translocation  
* pit_tag_ref: pit tag id  
* sex: m or f, re-coded as indicator variable (variable name = "male"): 1 = male, 0 = female   
* length: frog snout-vent length at capture at donor site (in mm)  
* weight: frog weight at capture at donor site (in grams)  
* bd_load: number of ITS copies at release, measured from swab collected at capture at donor site  
* survival: individual-level estimated probability of survival 1 year following translocation, based on cmr surveys conducted for at least two years post-translocation. Re-coded as indicator variable (variable name = "surv"): 0 = survival probability < 0.5; 1 = survival probability $\geq$ 0.5)

Other variables
* Maximum depth and surface area would seem useful to include, but 74976 is a stream/meadow habitat and as such depth and area at that site are likely not comparable to depth and area at lake habitats.  
* Frog condition is problematic because size and residuals from length:weight regression line are correlated. Specifically, the largest frogs have the largest positive residuals from a length:weight regression line, even after log10 transformation. As such, length and condition are not independent.

```{r frog-length-weight-plot}
read_csv(here::here("data", "clean", "frog_translocation_final.csv")) %>% 
ggplot(aes(x = length, y = weight)) +
  geom_point() +
  geom_smooth(method = "lm") +
  scale_x_log10() +
  scale_y_log10() +
  labs(x = "Length (mm)", 
       y = "Weight (g)")
```

* Another consideration when designing the analysis is that there are 12 records in which bd_load is null. Of these, 9 are from 70413 on 2013-07-15, likely due to a rush to get frogs packaged prior to helicopter arrival. 39 frogs were translocated on that date, so plenty of swab results are available to represent that site. Prior to running the analyses, should drop the 12 records where bd_load = NA. If bd_load is not an important predictor, can run analysis (without the bd_load predictor) using the full dataset.

## Exploratory analyses

Exploratory data analyses conducted in translocation_survival_exploratory.Rmd provided several notable insights.

### Frog survival across recipient sites
* 1-year frog survival is highly variable across sites, but highly repeatable within a site. This indicates the importance of including site_id or translocation_id as group-level effects. It remains unclear whether including a translocation_id nested within site_id should be considered (or if this nested group-level effect is even feasible given the relatively small number of translocations per site).

### Donor site
* Donor site has a strong association with frog survival, with survival at 70459 > 70567 > 72996. Explored possible effects of bd_load and frog size as drivers of this pattern. Differences in survival between donor sites seem unrelated to bd_load, but appear closely associated with frog size. Whether frog size is the sole driver or is a result of frog size plus other unmeasured donor site attributes (e.g., degree of resistance against Bd infection) is unknown.  
  * Minimal differences between donor sites in median bd_load, but slightly higher at 72996. No obvious effect of bd_load on 1-year frog survival.  
  * Strong effect of frog size on 1-year survival, with frogs > 55 mm having substantially higher 1-year survival. Frogs >60 mm had by far the highest survival.  
  * The effect of frog size on survival appears unrelated to bd_load because there is no relationship between frog size and bd_load.  
  * Frog size differs between donor site: size at 72996 < 70459 = 70567. This partially mirrors the observed differences in survival between sites. Sufficient overlap in frog sizes between donor sites that it might be possible to include both frog length and donor in the model.

### Bd load
* Frogs with the highest loads at the time of translocation had the lowest survival. The effect was only evident in the category that included frogs with loads of 56,000--2,100,000 ITS copies. Notable result given that the threshold level of 56,000 copies is /~ 1 order of magnitude below the 600,000 copies level associated with frog mortality during epizootics.

### Recipient site characteristics
* **Shore habitat**: The presence of shore habitat is associated with increased frog survival. Unknown if this is a direct effect reflecting the importance of this habitat type, or if shore habitat is serving as a proxy for another habitat feature. Tight association between shore habitat presence/absence and elevation will make it impossible to disentangle these effects, include only elevation in the model.  
* **Elevation**: Strong positive effect on frog survival.

### Translocation timing
* Day of the year showed no obvious relationship to frog survival.  
* Year was positively associated with survival. This is likely not a direct effect of year, but instead reflects better-informed selection of recipient sites with high-quality frog habitat over time. As such, consider not including year in the model.

### Winter severity
* No obvious relationship between either winter severity in year of translocation or in year following translocation on frog survival, but some substantial differences in frog survival between winter severity categories.

### Translocation order
* Frogs in the first cohort translocated to a site generally had higher survival than frogs in subsequent cohorts. However, effect appears relatively weak.  
* It remains unclear what mechanism could lead to the slightly higher mortality of later cohorts.

### Frog characteristics
* Frog size had strong effect on 1-year survival.  
* Minimal differences in survival between males and females (slightly higher in males).

@@

As originally conceived: survival \~ elevation + shore + snow_t + snow_t1 + year + day + order + donor + sex + length + log(bdload) \| site_id

Based on exploratory analyses: survival \~ elevation + snow_t + snow_t1 + day + order + donor + sex + length + log(bdload) \| site_id

Pairs plot to identify correlations between predictors (ggpairs from GGally package):
* It may not be possible to include both donor and length in a single model because the distribution of frog lengths is quite different across donor sites, causing dependency between donor and length. As such, these two variables have the potential to compete for the same variation. Need to look at this more closely, perhaps by plotting frog lengths at each donor site. Also, plot length against survival, grouped by donor.  Are these distributions as donor-specific as they appear to be? Because of the importance of distinguishing frog size effects from other effects associated with donor populations (e.g., inherent differences in Bd susceptibility), may need to explore whether it would be possible to compare the fit of two models, each containing one of these two variables.  
* Also need to disentangle elevation from site_id - currently only one elevation per site_id. Try breaking continuous elevation into categories, with > 2 site_ids per category. Hopefully, that will make elevation variable less dependent on site_id. A similar situation may exist for the two winter severity variables and site_id.  
* Shore and elevation may also be closely associated. Are all sites with shore = 0 at low elevation? If so, may need to only include elevation in model.

Group-level effects: 
The dataset has three sources of hierarchical structure: site_id, translocation_id, and donor site. Donor site is not suitable as a grouping variable because there are only 3 levels (i.e., site ids), not enough to make this useful. Site_id and translocation_id contain 12 and 24 levels, respectively, making them suitable as grouping variables. Including both site_id and translocation_id as nested grouping variables (i.e., translocations nested within site ids) also seems justified. Having more levels within a group would seem to increase the power to detect effects of predictors. With fewer levels/less variability, effect may be accounted for primarily by grouping variable itself instead of an associated predictor (e.g., elevation is closely associated with both site_id and translocation_id). With only 1-4 translocations per site and only 12 site ids total, it is possible that a model with site_id as sole grouping variable or a model with site_id and translocation_id as nested grouping variables will both attribute more of the variation in frog survival to the grouping variables and less to the predictors. A priori, that would suggest that the model with translocation_id as the grouping variable might be the most useful model structure (assuming that all models have similar posterior predictive accuracy). Lacking any strong rationale for one model structure over another, build 3 models and compare their posterior predictive accuracy using loo.

Interactions between predictor variables: General guidance is that interactions between the most important predictors should be included in model.

## Final steps before building model (in analysis Rmd):

* Bin elevation into "low" and "high" groups.  
* Recode order as 0/1 (instead of 1--4): "first".  
* Recode survival as 0/1 (instead of continuous probability): "surv".  
* Recode binary variables as 0/1 indicators: elevation ("elev_lo"), sex ("male"). See Gelman reference below).  
* Format relevant variables as factors: surv, elev_lo, shore, first, donor, male, year, site_id, translocation_id.  
* Standardize continuous variables: elevation, snow_t, snow_t1, day, length. Possibly standardize log(bd_load) also, but see Gelman for recommendations. Gelman recommends centering by subtracting the mean and standardizing by dividing by 2 standard deviations. Centering to a mean of 0 makes interactions more interpretable. Standardizing allows assessment of importance of predictors by comparing regression coefficients. Dividing by 2 SD puts continuous variables on approximately the same scale as binary (0/1) variables. This is true only if binary predictors have similar number of 0s and 1s. In the current dataset, there is some deviation from 1:1. Sex: female = 428, male = 351; order: 0 = 507, 1 = 272. Binary variables should be centered but not standardized. Variables in need of log-transformation should be transformed and then rescaled, although rescaling will not always be necessary.

Relevant references:  
* Gelman 2008: https://doi.org/10.1002/sim.3107  
* Schielzeth 2010: https://doi.org/10.1111/j.2041-210X.2010.00012.x

## Priors
* For regression coefficients in logistic regression: A recommended weakly informative prior is beta ~ student_t(nu,0,s), where s is chosen to provide weak information on the expected scale, and 3<nu<7.  https://github.com/stan-dev/stan/wiki/Prior-Choice-Recommendations.  
* Other recommendations in:  
  * Moving beyond noninformative priors: why and how to choose weakly informative priors in Bayesian analyses: https://doi.org/10.1111/oik.05985.  
  * Choosing priors in Bayesian ecological models by simulating from the prior predictive distribution: https://doi.org/10.1002/ecs2.3739.

## Next steps
* Frog size had a marginally important effect on frog survival. Given association between frog size and donor varibles, create a second model that includes only donor = 72996. Given that there is considerable variation in frog sizes translocated from this site, this might allow an assessment of the effect of frog size independent of any effect of donor.  
* To assess within-site association between survival for the first and subsequent translocated cohorts, build another dataset that includes all recipient sites for which > 1 translocation was conducted. Include survival from the first translocation as the response variable and survival from the subsequent translocations as a predictor variable. Other predictors could also be included.
